---
title: "BUNMD Supplementary Geography Variables File"
date: "2023-02-14"
output:
  pdf_document: default
  html_document: default
---

To create this file, we will need to read in a truncated version of the BUNMD and ZIP code of death variables files along withplace of birth variables that were created using Seth Sanders' Crosswalk.

To construct the “place of birth” variables, we used a crosswalk originally developed for this paper: 

Black, Dan A., Seth G. Sanders, Evan J. Taylor, and Lowell J. Taylor. 2015. “The Impact of the Great Migration on Mortality of African Americans: Evidence from the Deep South.” American Economic Review 105(2):477–503. doi: 10.1257/aer.20120642.

```{r}
library(tidyverse)
library(data.table)
library(dplyr)
```

Here, I read in the data and select only for the variables of interest.
```{r}
Zip_Features <- read.csv("Zip_Features.csv", colClasses = c(zip = "character",FIPS = "character")) 

Zip_Features <- subset(Zip_Features, select=-c(RUC1974, area_name, conterminous, acceptable_cities,district_name,RUC1983,timezone,area_codes,OASDI_benefits_2005,OASDI_total_2005,type,elevation,state,delv_address,OASDI_average_2005,OASDI_disabled_2005,OASDI_benefits_2005,OASDI_retired_2005,geocode))

birthplaces_df <- fread("bunmd_Birthplace.csv", select=c("ssn","bpl_city","feature_id","county_name","Region"))

bunmd_trunc <- fread("bunmd_v2.csv", select=c("ssn", "zip_residence"))
```
Next, I rename the ZIP code death place varirables so that it'll be clear that these refer to places of death in the final file.
```{r}
Zip_Features <- Zip_Features %>% 
  rename(
    death_fips = names(Zip_Features)[1],
    death_zip = names(Zip_Features)[2],
    death_state = names(Zip_Features)[3],
    death_city = names(Zip_Features)[4],
    death_county = names(Zip_Features)[5],
    death_country = names(Zip_Features)[6],
    death_region = names(Zip_Features)[7],
    death_ruc1993 = names(Zip_Features)[8]
  )
head(Zip_Features)
```
I truncate the BUNMD zip_residence variable so that it only has 5 digits. This is an important step since the places of death file only contains 5 digit ZIP codes, and gives us a variable that we can match on. 
```{r}
bunmd_trunc <- bunmd_trunc %>% 
  mutate(
    zip = str_trunc(as.character(zip_residence), width = 5, side = "right", ellipsis = "")
  ) %>% 
  select(-2) %>% 
  rename(death_zip = names(.)[2])

head(bunmd_trunc)
```
Once we have the matching ZIP code variables, we merge the information from the ZIP code file to the BUNMD file. This results in a new file that contains a person's SSN, ZIP code of death, and additional information sourced from the ZIP code file, such as the corresponding city, county, and state.
```{r}
bunmd_trunc <- merge(bunmd_trunc, Zip_Features, by = "death_zip", all.x = TRUE,sort = FALSE)
rm(Zip_Features)
head(bunmd_trunc)
```
Next, we rename birth place variables.
```{r}
birthplaces_df <- birthplaces_df %>% 
  rename(
    birth_city = names(birthplaces_df)[2],
    birth_gnis_code = names(birthplaces_df)[3],
    birth_county = names(birthplaces_df)[4],
    birth_region = names(birthplaces_df)[5])
```
Next, we merge birth place variables to the BUNMD file. These variables include the city, county, and state where the person was born. Now, we have a new file that contains information on people's death and birthplaces, in addition to their SSN.
```{r}
birthplaces_df <- merge(bunmd_trunc, birthplaces_df, by = "ssn", all.x = TRUE,sort = FALSE)
rm(bunmd_trunc)
head(birthplaces_df)
```
In the chunks of code below, we standardize the data set and make everything lowercase. We also make sure that any differences in wording are standardized. 
```{r}
birthplaces_df$birth_city <- tolower(birthplaces_df$birth_city)
birthplaces_df$birth_county <- tolower(birthplaces_df$birth_county)
birthplaces_df$birth_region <- tolower(birthplaces_df$birth_region)
birthplaces_df$death_state <- tolower(birthplaces_df$death_state)
birthplaces_df$death_city <- tolower(birthplaces_df$death_city)
birthplaces_df$death_county <- tolower(birthplaces_df$death_county)
head(birthplaces_df)
```
```{r}
birthplaces_df <- birthplaces_df %>% 
  mutate(
    death_region = ifelse(death_region == "North_East", "northeast", death_region)
  )
birthplaces_df$death_region <- tolower(birthplaces_df$death_region)
table(birthplaces_df$death_region)
```
```{r}
birthplaces_df <- birthplaces_df %>% 
  mutate(
    death_county = str_remove_all(death_county, "county"),
    death_county = str_trim(death_county)
  )
head(birthplaces_df$death_county, 10) #we remove the word "county" 
```
```{r}
birthplaces_df$death_state <- gsub("\\b\\s+", "", birthplaces_df$death_state)

birthplaces_df <- birthplaces_df %>% 
  mutate(
    death_state = case_when(
      death_state == "" ~ NA_character_,
      death_state == "arizona14/       " ~ "arizona",
      death_state == "newhampshire" ~ "new hampshire",
      death_state == "southcarolina" ~ "south carolina",
      death_state == "newjersey" ~ "new jersey",
      death_state == "newmexico" ~ "new mexico",
      death_state == "northdakota" ~ "north dakota",
      death_state == "northcarolina" ~ "north carolina",
      death_state == "southdakota" ~ "south dakota",
      death_state == "westvirginia" ~ "west virginia",
      death_state == "rhodeisland" ~ "rhode island",
      TRUE ~ death_state
    )
  )
unique(birthplaces_df$death_state)
```
Then, we remove all people who have neither a birth or a death city. 
```{r}
birthplaces_df <- birthplaces_df %>% 
  mutate(
    deletes = paste(birth_city, death_city)
  ) %>% 
  filter(!deletes %in% "NA NA") %>% 
  select(-deletes)
head(birthplaces_df)
```
We then make sure that the FIPS and ZIP code variables are coded as string variables so that the leading zeroes do not get truncated.  
```{r}
birthplaces_df$death_zip <- as.character(birthplaces_df$death_zip)
birthplaces_df$death_county_fips <- as.character(birthplaces_df$death_county_fips)
```
Lastly, we save the file as a csv
```{r}
write.csv(birthplaces_df,"birth_death_features.csv",row.names = FALSE)
```


