---
title: "BUNMD Supplementary Geography Variables File Codebook"
output:
  pdf_document:
header-includes:
- \usepackage{hyperref}
- \usepackage{float}
- \restylefloat{table}
- \usepackage{caption}
---
[^updated]: Last updated: `r format(Sys.time(), '%d %B, %Y')`



| Page| Variable            | Label                                        |
|----:|:--------------------|:---------------------------------------------|
|  \hyperlink{page.2}{2}    | \hyperlink{page.2}{ssn}                |Social Security Number                        |
|  \hyperlink{page.3}{3}    | \hyperlink{page.3}{birth\_city}              |City of Birth                           |
|  \hyperlink{page.4}{4}    | \hyperlink{page.4}{birth\_gnis\_code}              |GNIS Code of Birth Place            | 
|  \hyperlink{page.5}{5}    | \hyperlink{page.5}{birth\_county}              |County Name of Birth                    |
|  \hyperlink{page.6}{6}    | \hyperlink{page.6}{birth\_region}                |Census Region of Birth                |
|  \hyperlink{page.7}{7}    | \hyperlink{page.7}{death\_zip}         |ZIP Code of Death                |
|  \hyperlink{page.8}{8}    | \hyperlink{page.8}{death\_state}   |FIPS code of Death County                |
|  \hyperlink{page.9}{9}    | \hyperlink{page.9}{death\_state}  |State of Death                        |
|  \hyperlink{page.10}{10}  | \hyperlink{page.10}{death\_city}         |Primary ZIP City of Death                  |
|  \hyperlink{page.11}{11}  | \hyperlink{page.11}{death\_county}   |ZIP County of Death                |
|  \hyperlink{page.12}{12}  | \hyperlink{page.12}{death\_country}  |Country of Death                            |
|  \hyperlink{page.13}{13}  | \hyperlink{page.13}{death\_region}       |Death Census Region                |
|  \hyperlink{page.14}{14}  | \hyperlink{page.14}{death\_ruc1993}       |USDA Rural-Urban Continuum Code              |


**Summary:**  The Berkeley Unified Numident Mortality Detailed Supplementary Geography Variables File (N = 42,878,885) provides information on a person's birth and death locations. This dataset provides details on the death location based on the deceased's ZIP code. It consists of 29,676,585 entries specifying birthplaces at city, county, and state levels, but it excludes data on individuals born in Washington D.C. or U.S. territories. City and county information is derived from Seth Sanders' crosswalk, while state-level data originates from the BUNMD itself.

To combine the Berkeley Unified Numident Mortality Crosswalk Database with the BUNMD, use the individual's social security number as the linking variable.
\newpage 



```{r echo = F, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(data.table)
library(tidyverse)

BUNMD_Geography_File <- fread("BUNMD_Geography_File.csv")
```

\newpage 

**\huge ssn**
\normalsize
\vspace{12pt}

**Label**: Social Security Number

*Description*: ssn reports a person’s Social Security number as recorded in the Numident death records.
This variable uniquely identifies all records in the dataset and can be used to merge onto the BUNMD dataset.


\newpage 

**\huge birth_city **
\normalsize
\vspace{12pt}

**Label**: City of birth

*Description*: birth_city reports the city of birth recorded in the BUNMD database. This variable is the result of the linkage between the BUNMD and Crosswalk databases. Approximately 94% of individuals in the BUNMD with city of birth information and non-missing feature_id were matched with the crosswalk (29,676,585). 

\newpage 


**\huge birth_gnis_code **
\normalsize
\vspace{12pt}

**Label:** birth_gnis_code

*Description*: birth_gnis_code is a numeric variable that denotes an individuals’ Geographic Names Information System (GNIS) code for their place of birth. The GNIS code maps onto physical locations located in the U.S., including longitude and latitude coordinates and physical feature names. 

For more information on GNIS codes, please see <https://www.usgs.gov/faqs/what-geographic-names-information-system-gnis>.


\newpage 


**\huge birth_county **
\normalsize
\vspace{12pt}

**Label**: County name

*Description*: birth_county reports the county of birth for an individual as a character string. 

\newpage 


**\huge birth_region **
\normalsize
\vspace{12pt}

**Label**: Census Region

*Description*:  birth_region is a categorical variable that categorizes an individual’s place of birth into one of the four Census Regions – Northeast, Midwest, South, and West – based on their state of birth. 

For more information on Census Regions, please see <https://www.census.gov/programs-surveys/economic-census/guidance-geographies/levels.html>.


```{r echo = F, message=FALSE, warning=FALSE, fig.pos="H"}
library(ggplot2)

# Filter out NA values from birth_region column
birth_region_no_na <- na.omit(BUNMD_Geography_File$birth_region)

# Create plot
ggplot(data = data.frame(birth_region = birth_region_no_na), aes(x = birth_region)) +
  geom_bar(fill = "lightblue", color = "white") +
  ggtitle("Frequency of Values") +
  xlab("Census Region of Birth") +
  ylab("Frequency") +
  scale_y_continuous(limits = c(0, max(table(birth_region_no_na)) + 1), expand = c(0,0), labels = scales::comma) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line.x = element_line(color = "black", size = 0.5),
        axis.line.y = element_line(color = "black", size = 0.5),
        axis.title = element_text(size = 12),
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5))
```

\newpage 

**\huge death_zip **
\normalsize
\vspace{12pt}

**Label**: Death ZIP Code

*Description*: death_zip is the ZIP code that is recorded at the time of death in the BUNMD. ZIP codes are primarily used for USPS mail delivery purposes, not as geographic units.

For a more detailed description of ZIP codes, please see: <https://faq.usps.com/s/article/ZIP-Code-The-Basics>.

\newpage 

**\huge death_fips **
\normalsize
\vspace{12pt}

**Label**: Death County FIPS Code

*Description*: death_fips, a numeric variable, is the FIPS code for an individual’s county of death as sourced from their ZIP code of death. 

For a more detailed description of FIPS codes, please see: <https://transition.fcc.gov/oet/info/maps/census/fips/fips.txt#:~:text=FIPS%20codes%20are%20numbers%20which,to%20which%20the%20county%20belongs>.

\newpage 

**\huge death_state **
\normalsize
\vspace{12pt}

**Label**: State of Death

*Description*: death_state reports on the state the death ZIP code falls into. 

For a more detailed description of ZIP codes, please see: <https://faq.usps.com/s/article/ZIP-Code-The-Basics>.

\newpage 

**\huge death_city **
\normalsize
\vspace{12pt}

**Label**: City of Death

*Description*: death_city reports a ZIP code's primary city according to the USPS (the city of primary USPS office for the ZIP code). In the case that a ZIP code crosses a city line, the city listed will be where the primary post office for that ZIP code is located.  

For a more detailed description of ZIP codes, please see: <https://faq.usps.com/s/article/ZIP-Code-The-Basics>.

\newpage 

**\huge death_county **
\normalsize
\vspace{12pt}

**Label**: County of Death

*Description*: death_county reports a ZIP code's county
according to the USPS (the city of primary USPS office for the ZIP
code). 

\newpage 

**\huge death_country **
\normalsize
\vspace{12pt}

**Label**: Country of Death

*Description*: death_country indicates the country where the reported ZIP code is located. While the majority of these ZIP codes are within the United States, countries that host U.S. military bases on their territories may also have ZIP codes associated with them.

\newpage 

**\huge death_region **
\normalsize
\vspace{12pt}

**Label**: Census Region of Death

*Description*: death_region is a categorical variable that categorizes an individual’s place of death, based on their ZIP code and state of death, into one of the four Census Regions – Northeast, the Midwest, the South, and the West. 

\newpage 

**\huge death_ruc1993 **
\normalsize
\vspace{12pt}

**Label**: USDA Rural-Urban Continuum Code

*Description*: death_ruc1993 is sourced from the United States Department of Agriculture (USDA) Urban-Rural Continuum Codes, which are a classification system developed to categorize U.S. counties based on their degree of urbanization and adjacency to metropolitan areas. The system consists of ten codes, ranging from 0 to 9, where lower codes represent more urbanized counties and higher codes represent more rural areas.

For more information: <https://www.ers.usda.gov/data-products/rural-urban-continuum-codes.aspx>